---
title: "PM566 Final Project"
author: "Haoyue Shan"
output: 
    html_document:
        toc: TRUE
        toc_float: TRUE
---

This is my PM566 Final Project website.

```{r setup, message=FALSE, echo=FALSE, warning=FALSE}

library(data.table)
library(tidyverse)
library(dplyr)
library(plotly)
library(DT)
library(knitr)

# Initialize code chunk options
opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  eval=TRUE,
  echo = FALSE,
  cache = FALSE,
  fig.width = 7, 
  fig.align = 'center',
  fig.asp = 0.618,
  out.width = "700px",
  class.source = "code-r")
```

```{css, echo = FALSE}
.code-r { /* Code block */
  font-size: 15px;
}

.code-r-small { /* Code block */
  font-size: 10px;
}
```
```{r}
source("process_COVID_data.R")
```

<br>

## lab 12 
You need to name each code chunk, like this one: `plot1`
```{r plot1, class.source="code-r-small"}

p1_scatter <- cv_states_today %>% 
  plot_ly(x = ~pop_density, y = ~deathsper100k,
          type = 'scatter', mode = 'markers', color = ~state,
          size = ~population, sizes = c(5, 70), marker = list(sizemode='diameter', opacity=0.5),
          hoverinfo = 'text',
          text = ~paste( paste(state, ":", sep=""), paste(" Cases per 100k: ", per100k, sep="") , paste(" Deaths per 100k: ",
                        deathsper100k, sep=""), sep = "<br>")) %>%
  layout(title = "Population-normalized COVID-19 deaths vs. population density",
                  yaxis = list(title = "Deaths per 100k"), xaxis = list(title = "Population Density"),
         hovermode = "compare")

# filter out "District of Columbia"
cv_states_today_scatter <- cv_states_today %>% filter(state!="District of Columbia")

p2_scatter <- cv_states_today_scatter %>% 
  plot_ly(x = ~pop_density, y = ~deathsper100k,
          type = 'scatter', mode = 'markers', color = ~state,
          size = ~population, sizes = c(5, 70), marker = list(sizemode='diameter', opacity=0.5),
          hoverinfo = 'text',
          text = ~paste( paste(state, ":", sep=""), paste(" Cases per 100k: ", per100k, sep="") , paste(" Deaths per 100k: ",
                        deathsper100k, sep=""), sep = "<br>")) %>%
  layout(title = "Population-normalized COVID-19 deaths vs. population density",
                  yaxis = list(title = "Deaths per 100k"), xaxis = list(title = "Population Density"),
         hovermode = "compare")
```

<br>

## showcase {.tabset}

Create tabs to display each figure

### Scatterplot: with DC

```{r p1}
p1_scatter
```

### Scatterplot: without DC

```{r p2}
p2_scatter
```

## {-}






# For Assignment 5
```{r,message=FALSE, echo=FALSE, warning=FALSE, results='hide',fig.show='hide'}
# data manipulation
library(data.table)
library(tidyverse) 
library(psych)
library(leaflet)
library(ggplot2)
library(kableExtra)
library(plotly)
#look at the data
covid_race = data.table::fread("/Users/haoyueshan/Desktop/pm566/final-redo/data/Race-Data-updated.csv")
#choosing the latest data 
covid_race = covid_race[1:56,1:28]

#change characteristic varibles into numeric varibles
covid_race = covid_race %>% 
  mutate(Cases_Total = as.numeric(Cases_Total), Cases_White = as.numeric(Cases_White), Cases_Ethnicity_Unknown = as.numeric(Cases_Ethnicity_Unknown))
sub = covid_race[,4:28]

#sum up the total values without NA's for every race and make an overall barplot
temp = apply(sub,2,sum, na.rm = TRUE)
temp = as.data.frame(temp)
names=rownames(temp)

#make subgroup of cases and deaths(only look at the known cases and deaths)
cases_name = names[0:8]
cases_number= temp[0:8,]
cases = data.frame(cases_name,cases_number)
deaths_name = names[14:21]
deaths_number = temp[14:21,]
deaths = data.frame(deaths_name,deaths_number)
kable(cases,'html', caption = "The total cases number of known races")
kable(deaths,'html', caption = "The total deaths number of known races")
#count the percentage for each group 
pct_case = cases$cases_number/sum(cases$cases_number)
pct_cases = data.frame(cases_name,pct_case)
kable(pct_cases,'html', caption = "The percentage of cases for the known races")
pct_death = deaths$deaths_number/sum(deaths$deaths_number)
pct_deaths = data.frame(deaths_name,pct_death)
kable(pct_deaths,'html', caption = "The percentage of deaths for the known races")

#making a pie chart for race compnent using plotly 

cases = plot_ly(pct_cases, labels = ~cases_name, values = ~pct_case, type = 'pie')
cases = cases %>% layout(title = 'The percentage of reported COVID cases for each known races',
         xaxis = list(showgrid = FALSE, zeroline = FALSE, showticklabels = FALSE),
         yaxis = list(showgrid = FALSE, zeroline = FALSE, showticklabels = FALSE))

deaths = plot_ly(pct_deaths, labels = ~deaths_name, values = ~pct_death, type = 'pie')
deaths = deaths %>% layout(title = 'The percentage of reported COVID deaths for each known races',
         xaxis = list(showgrid = FALSE, zeroline = FALSE, showticklabels = FALSE),
         yaxis = list(showgrid = FALSE, zeroline = FALSE, showticklabels = FALSE))

#bar plot for each table using plotly
bar_cases = plot_ly(
  x = cases_name,
  y = pct_case,
  name = "Reported COVID cases",
  type = "bar"
)

bar_deaths = plot_ly(
  x = deaths_name,
  y = pct_death,
  name = "Reported COVID deaths",
  type = "bar"
)


##the fatality rate and plot
races = c("White","Black","LatinX","Asian","AIAN","NHPI","Cases_Multiracial","Cases_Other")
total_number = data.frame(races,cases_number,deaths_number)
total_number = total_number %>%
  mutate(fatality_rate = deaths_number/cases_number)
kable(total_number,'html', caption = "The case number, death number and fatality rate for each race")

fig2 <- plot_ly(total_number, x = ~fatality_rate, y = ~races, type = 'bar', orientation = 'h', marker = list(color = 'rgba(246, 78, 139, 0.6)'))
```

## Graphs {.tabset}

### Pie chart of cases number for known races 
```{r cases}
cases
```
From the pie chart we can see that the white race contributed more than a half(51.1%) to the total number of reported COVID cases.Other races made up the other half of this pie chart. We can see that besides white race, the black race contributed most(20.1%) to the pie chart; the third large component is latino race, which contribute 15.8% to the total cases number; the forth large component is the asian race, which contributed less than 10%(7.92%) to the total cases number. Besides those races mentioned above, the AIAN, NHPI, multiracial and other races added up contributed less than 10% to the total cases number. 

### Pie chart of deaths number for known races
```{r deaths}
deaths
```

From the pie chart above, we can tell that race white contributed more than a half to the death counts(52.7%), the race black(19.9%) and race latino(19%) contributed similar porportion to the total death count. As for other races, they added up contributed less than 10% to the total death number. 


### Bar chart of cases number for known races
```{r bar_cases}
bar_cases
```
From the bar plot above, we can see that the white race have the highest cases number among all of the known races, which is more than twice as much as the second highest race latino.  Besides white, latino and black, each of the other races contribute less than 10% to the total number.

### Bar chart of death number for known races
```{r bar_deaths}
bar_deaths
```
From the bar chart above, we can see that the white race have the highest death number among all of the known races, and the proportion of race black is slightly higher than race latino. Besides race white, black and latino, each of the other races contributed less than 5% to the total death number.

### Fatality rate for know races 
```{r fig2}
fig2
```
From the table and bar plot above, we can see that though race asian has the fifth largest number of cases count and forth largest number of death count, it has the highest mortality rate among all of the known races.

## {-}

<br>



<br>
<br>

